Formalizing the Kolmogorov extension Theorem in Lean

### 2024-07-22 Manuscript/proof_outline.tex
use IsSetSemiring from mathlib, lint

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
structure SetSemiring (C : Set (Set α))
    : Prop where
  empty_mem : ∅ ∈ C
  inter_mem : ∀ (s) (_ : s ∈ C) (t) (_ : t ∈ C), s ∩ t ∈ C
  diff_eq_Union' :
    ∀ (s) (_ : s ∈ C) (t) (_ : t ∈ C),
    ∃ (I : Finset (Set α)) (_h_ss : ↑I ⊆ C)
    (_h_dis : PairwiseDisjoint (I : Set (Set α)) id),
      t \ s = ⋃₀ I
\end{minted}

@@

\begin{minted}[highlightlines={1-4}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem setField_cylinders : SetField (cylinders α) 
theorem setRing_cylinders :  SetRing (cylinders α)
theorem setSemiringCylinders
  : SetSemiring (cylinders α)   
\end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem exists_disjoint_finset_diff_eq
    (hC : SetSemiring C) (hs : s ∈ C)
    (I : Finset (Set α)) (hI : ↑I ⊆ C) :
    ∃ (J : Finset (Set α)) (_h_ss : ↑J ⊆ C)
    (_h_dis : PairwiseDisjoint (J : Set (Set α)) id),
    s \ ⋃₀ I = ⋃₀ J
\end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def diff₀ (hC : SetSemiring C)
    (hs : s ∈ C) (I : Finset (Set α)) (hI : ↑I ⊆ C)
    [DecidableEq (Set α)] : Finset (Set α) :=
  (hC.exists_disjoint_finset_diff_eq hs I hI).choose \ {∅}
\end{minted}

@@

%% \begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
%% def indexedDiff₀ (hC : SetSemiring C)
%%     (J : Finset (Set α)) (hJ : ↑J ⊆ C)
%%     (n : Fin J.card) : Finset (Set α) :=
%%   hC.diff₀ (ordered_mem' hJ n) (finsetLT J n)
%%     (finsetLT_subset' J hJ n)
%% \end{minted}

@@

def extendContent
    (hC : SetSemiring C)
    (m : ∀ s : Set α, s ∈ C → ℝ≥0∞)
    (m_empty : m ∅ hC.empty_mem = 0)
    (m_add :
      ∀ (I : Finset (Set α)) (h_ss : ↑I ⊆ C)
      (_h_dis : PairwiseDisjoint (I : Set (Set α)) id)
      (h_mem : ⋃₀ ↑I ∈ C),
      m (⋃₀ I) h_mem = ∑ u : I, m u (h_ss u.prop)) :
    AddContent C 
\end{minted}

@@

def kolContent
    (hP : IsProjectiveMeasureFamily P) :
    AddContent (cylinders α) :=
  extendContent setSemiringCylinders
    (kolmogorovFun P) (kolmogorovFun_empty hP)
    (kolmogorovFun_additive hP)
\end{minted}

@@

%% \begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
%% def ofAddSubaddCaratheodory
%%     (hC : SetSemiring C)
%%     (m : ∀ s : Set α, s ∈ C → ℝ≥0∞)
%%     (m_empty : m ∅ hC.empty_mem = 0)
%%     (m_add : ∀ (I : Finset (Set α)) (h_ss : ↑I ⊆ C)
%%       (_h_dis : PairwiseDisjoint (I : Set (Set α)) id)
%%       (h_mem : ⋃₀ ↑I ∈ C),
%%       m (⋃₀ I) h_mem = ∑ u : I, m u (h_ss u.prop))
%%     (m_sigma_subadd : ∀ ⦃f : ℕ → Set α⦄
%%       (hf : ∀ i, f i ∈ C) (hf_Union : (⋃ i, f i) ∈ C),
%%       m (⋃ i, f i) hf_Union ≤ ∑' i, m (f i) (hf i)) :
%%     @Measure α (inducedOuterMeasure m
%%       hC.empty_mem m_empty).caratheodory
%% \end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def Measure.ofAddContent
    [mα : MeasurableSpace α] (hC : SetSemiring C)
    (hC_gen : MeasurableSpace.generateFrom C = mα)
    (m : AddContent C)
    (m_sigma_subadd : ∀ ⦃f : ℕ → Set α⦄
      (hf : ∀ i, f i ∈ C) (hf_Union : (⋃ i, f i) ∈ C),
      m (⋃ i, f i) hf_Union ≤ ∑' i, m (f i) (hf i)) :
    Measure α
\end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem sigma_additive_addContent_of_tendsto_zero
    (hC : SetRing C)
    (m : AddContent C)
    (hm_ne_top : ∀ {s} (_ : s ∈ C), m s ≠ ∞)
    (hm : ∀ ⦃s : ℕ → Set α⦄ (_ : ∀ n, s n ∈ C),
      Antitone s → (⋂ n, s n) = ∅ →
      Tendsto (fun n => m (s n)) atTop (nhds 0))
      ⦃f : ℕ → Set α⦄ (hf : ∀ i, f i ∈ C) (hUf : (⋃ i, f i) ∈ C)
      (h_disj : Pairwise (Disjoint on f)) :
    m (⋃ i, f i) = ∑' i, m (f i) 
\end{minted}

@@

%% lemma AddContent.sigma_additive_of_regular [Nonempty α] (hR : SetRing R) (m : AddContent R)
%%     (hm_ne_top : ∀ {s} (_ : s ∈ R), m s ≠ ∞)
%%     (hC : IsCompactSystem C) (hCR : C ⊆ R)
%%     (h_reg : ∀ A (_ : A ∈ R) (ε : ℝ≥0∞) (_ : 0 < ε), ∃ K ∈ C, K ⊆ A ∧ m (A \ K) ≤ ε)
%%     ⦃f : ℕ → Set α⦄ (hf : ∀ i, f i ∈ R) (hUf : (⋃ i, f i) ∈ R) (h_disj : Pairwise (Disjoint on f)) :
%%     m (⋃ i, f i) = ∑' i, m (f i)

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def projectiveLimitWithWeakestHypotheses
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    (hP : IsProjectiveMeasureFamily P) :
    Measure (∀ i, α i) :=
  Measure.ofAddContent setSemiringCylinders
    generateFrom_cylinders (kolContent hP)
    (kolContent_sigma_subadditive_of_innerRegular hP
      fun J => innerRegular_isCompact_isClosed_
      measurableSet_of_complete_countable (P J))    
\end{minted}
\sloppy Let us look closer at the formalized proof:
\leanlinecolor{kolContent_sigma_subadditive_of_innerRegular} has the
same hypotheses as
\leanlinecolor{kolContent_sigma_additive_of_innerRegular},
except \leanline{h_disj}. So, the lemma in the last brackets shows
$\sigma$-sub-additivity of the content $P$ from
Theorem~\ref{T1}. Then, \leanlinecolor{generateFrom_cylinders} gives
the \leanline{MeasurableSpace} on which we define the
\leanlinecolor{addContent}.

>>>>

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
structure IsSetSemiring (C : Set (Set α))
    : Prop where
  empty_mem : ∅ ∈ C
  inter_mem : ∀ (s) (_ : s ∈ C) (t) (_ : t ∈ C), s ∩ t ∈ C
  diff_eq_Union' :
    ∀ (s) (_ : s ∈ C) (t) (_ : t ∈ C),
    ∃ (I : Finset (Set α)) (_h_ss : ↑I ⊆ C)
    (_h_dis : PairwiseDisjoint (I : Set (Set α)) id),
      t \ s = ⋃₀ I
\end{minted}

@@

\begin{minted}[highlightlines={1-4}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theoremIsSetField_cylinders :IsSetField (cylinders α) 
theorem setRing_cylinders :  IsSetRing (cylinders α)
theorem IsSetSemiringCylinders
  : IsSetSemiring (cylinders α)   
\end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem exists_disjoint_finset_diff_eq
    (hC : IsSetSemiring C) (hs : s ∈ C)
    (I : Finset (Set α)) (hI : ↑I ⊆ C) :
    ∃ (J : Finset (Set α)) (_h_ss : ↑J ⊆ C)
    (_h_dis : PairwiseDisjoint (J : Set (Set α)) id),
    s \ ⋃₀ I = ⋃₀ J
\end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def diff₀ (hC : IsSetSemiring C)
    (hs : s ∈ C) (I : Finset (Set α)) (hI : ↑I ⊆ C)
    [DecidableEq (Set α)] : Finset (Set α) :=
  (hC.exists_disjoint_finset_diff_eq hs I hI).choose \ {∅}
\end{minted}

@@

%% \begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
%% def indexedDiff₀ (hC : IsSetSemiring C)
%%     (J : Finset (Set α)) (hJ : ↑J ⊆ C)
%%     (n : Fin J.card) : Finset (Set α) :=
%%   hC.diff₀ (ordered_mem' hJ n) (finsetLT J n)
%%     (finsetLT_subset' J hJ n)
%% \end{minted}

@@

def extendContent
    (hC : IsSetSemiring C)
    (m : ∀ s : Set α, s ∈ C → ℝ≥0∞)
    (m_empty : m ∅ hC.empty_mem = 0)
    (m_add :
      ∀ (I : Finset (Set α)) (h_ss : ↑I ⊆ C)
      (_h_dis : PairwiseDisjoint (I : Set (Set α)) id)
      (h_mem : ⋃₀ ↑I ∈ C),
      m (⋃₀ I) h_mem = ∑ u : I, m u (h_ss u.prop)) :
    AddContent C 
\end{minted}

@@

def kolContent
    (hP : IsProjectiveMeasureFamily P) :
    AddContent (cylinders α) :=
  extendContent IsSetSemiringCylinders
    (kolmogorovFun P) (kolmogorovFun_empty hP)
    (kolmogorovFun_additive hP)
\end{minted}

@@

%% \begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
%% def ofAddSubaddCaratheodory
%%     (hC : IsSetSemiring C)
%%     (m : ∀ s : Set α, s ∈ C → ℝ≥0∞)
%%     (m_empty : m ∅ hC.empty_mem = 0)
%%     (m_add : ∀ (I : Finset (Set α)) (h_ss : ↑I ⊆ C)
%%       (_h_dis : PairwiseDisjoint (I : Set (Set α)) id)
%%       (h_mem : ⋃₀ ↑I ∈ C),
%%       m (⋃₀ I) h_mem = ∑ u : I, m u (h_ss u.prop))
%%     (m_sigma_subadd : ∀ ⦃f : ℕ → Set α⦄
%%       (hf : ∀ i, f i ∈ C) (hf_Union : (⋃ i, f i) ∈ C),
%%       m (⋃ i, f i) hf_Union ≤ ∑' i, m (f i) (hf i)) :
%%     @Measure α (inducedOuterMeasure m
%%       hC.empty_mem m_empty).caratheodory
%% \end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def Measure.ofAddContent
    [mα : MeasurableSpace α] (hC : IsSetSemiring C)
    (hC_gen : MeasurableSpace.generateFrom C = mα)
    (m : AddContent C)
    (m_sigma_subadd : ∀ ⦃f : ℕ → Set α⦄
      (hf : ∀ i, f i ∈ C) (hf_Union : (⋃ i, f i) ∈ C),
      m (⋃ i, f i) hf_Union ≤ ∑' i, m (f i) (hf i)) :
    Measure α
\end{minted}

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem sigma_additive_addContent_of_tendsto_zero
    (hC : IsSetRing C)
    (m : AddContent C)
    (hm_ne_top : ∀ {s} (_ : s ∈ C), m s ≠ ∞)
    (hm : ∀ ⦃s : ℕ → Set α⦄ (_ : ∀ n, s n ∈ C),
      Antitone s → (⋂ n, s n) = ∅ →
      Tendsto (fun n => m (s n)) atTop (nhds 0))
      ⦃f : ℕ → Set α⦄ (hf : ∀ i, f i ∈ C) (hUf : (⋃ i, f i) ∈ C)
      (h_disj : Pairwise (Disjoint on f)) :
    m (⋃ i, f i) = ∑' i, m (f i) 
\end{minted}

@@

%% lemma AddContent.sigma_additive_of_regular [Nonempty α] (hR : IsSetRing R) (m : AddContent R)
%%     (hm_ne_top : ∀ {s} (_ : s ∈ R), m s ≠ ∞)
%%     (hC : IsCompactSystem C) (hCR : C ⊆ R)
%%     (h_reg : ∀ A (_ : A ∈ R) (ε : ℝ≥0∞) (_ : 0 < ε), ∃ K ∈ C, K ⊆ A ∧ m (A \ K) ≤ ε)
%%     ⦃f : ℕ → Set α⦄ (hf : ∀ i, f i ∈ R) (hUf : (⋃ i, f i) ∈ R) (h_disj : Pairwise (Disjoint on f)) :
%%     m (⋃ i, f i) = ∑' i, m (f i)

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def projectiveLimitWithWeakestHypotheses
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    (hP : IsProjectiveMeasureFamily P) :
    Measure (∀ i, α i) :=
  Measure.ofAddContent IsSetSemiringCylinders
    generateFrom_cylinders (kolContent hP)
    (kolContent_sigma_subadditive_of_innerRegular hP
      fun J => innerRegular_isCompact_isClosed_
      measurableSet_of_complete_countable (P J))    
\end{minted}
\sloppy Let us look closer at the formalized proof:
\leanlinecolor{kolContent_sigma_subadditive_of_innerRegular} has the
same hypotheses as
\leanlinecolor{kolContent_sigma_additive_of_innerRegular},
except \leanline{h_disj}. So, the lemma in the last brackets shows
$\sigma$-sub-additivity of the content $P$ from
Theorem~\ref{T1}. Then, \leanlinecolor{generateFrom_cylinders} gives
the \leanline{MeasurableSpace} on which we define the
\leanlinecolor{addContent}.
